
---
title: "QA Report — SENEPA Tidy Dataset"
author: "SENEPA / PINV01-528 Team"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
  pdf_document:
    toc: true
    number_sections: true
fontsize: 11pt
lang: "en-GB"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(lubridate)
library(sf)
library(knitr)
library(ggplot2)
library(readr)
library(plotly)

if (!require('pacman'))
  install.packages('pacman',
                   repos = "http://cran.us.r-project.org"
  )

pacman::p_load(ENMeval,
               tidyverse,
               rio,
               raster,
               plotly,
               dismo,
               summarytools,
               explore,
               dlookr,
               leaflet)

```


### Executive summary

This report documents automated quality checks run against `data/processed_data/Dataset_Combinado.csv` (audited version). It quantifies common issues (invalid dates, coordinates outside Paraguay, exact duplicates, non-standard species names, missing critical fields), provides examples of problematic records, and produces reproducible artifacts (`reports/qa_summary.csv`, `logs/qa_examples.csv`, `logs/qa_report_metadata.csv`).

---

### 1 Methodology

- **Automated checks**: expected columns and types; date range (1998–2023); coordinates within Paraguay bounds; exact duplicates (SPECIES + DATE + LAT + LONG); species name sanity checks; NA rates for critical fields; coherence between `GEOREFERENCING_TYPE` and `GEOREFERENCING_UNCERTAINTY_M` when present.
- **Outputs**: summary tables, example records, plots (uncertainty histogram, spatial scatter), and metadata files for reproducibility.
- **Reproducibility**: the report writes `reports/qa_summary.csv`, `logs/qa_examples.csv`, and `logs/qa_report_metadata.csv`.

---

### 2 Load dataset and metadata

```{r load-data}
dataset_path <- "../data/processed_data/SENEPA_tidy_data_set.csv"
if (!file.exists(dataset_path)) stop("Dataset not found at: ", dataset_path)

dataset <- readr::read_csv(dataset_path, show_col_types = FALSE, locale = locale(encoding = "UTF-8"))
n_total <- nrow(dataset)
cat("Rows loaded:", n_total, "\n")
```

---

### 3 Checks and results

#### 3.1 Expected columns and basic types


```{r check-columns}
expected_cols <- c("DEPARTMENT","DISTRICT","LOCALITY","YEAR","DATE","DISEASE","SPECIES",
                   "UTM_ZONE","UTM_HEMISPHERE","UTM_X","UTM_Y","LAT","LONG","GEOREFERENCING_TYPE",
                   "GEOREFERENCING_UNCERTAINTY_M","DATA_SOURCE")
present_cols <- names(dataset)
cols_missing <- setdiff(expected_cols, present_cols)
cols_extra <- setdiff(present_cols, expected_cols)

kable(tibble(
  total_rows = n_total,
  cols_present = length(present_cols),
  cols_missing = paste(cols_missing, collapse = "; "),
  cols_extra = paste(cols_extra, collapse = "; ")
))
```

#### 3.2 Date parsing and range check

```{r check-dates}
dataset <- dataset %>%
  mutate(DATE2 = suppressWarnings(as.Date(DATE))) %>%
  mutate(YEAR2 = ifelse(!is.na(YEAR), as.numeric(YEAR), lubridate::year(DATE2)))

bad_dates <- dataset %>%
  filter((!is.na(DATE) & is.na(DATE2)) |
         (!is.na(DATE2) & (DATE2 < as.Date("1998-01-01") | DATE2 > as.Date("2023-12-31"))))

kable(tibble(
  total_rows = n_total,
  invalid_dates = nrow(bad_dates),
  pct_invalid_dates = round(100 * nrow(bad_dates) / n_total, 3)
))
```

#### 3.3 Coordinate bounds (Paraguay)

```{r check-coords}
lat_min <- -28; lat_max <- -19
long_min <- -63; long_max <- -54

dataset <- dataset %>%
  mutate(LAT = as.numeric(LAT), LONG = as.numeric(LONG))

bad_coords <- dataset %>%
  filter((!is.na(LAT) & (LAT < lat_min | LAT > lat_max)) |
         (!is.na(LONG) & (LONG < long_min | LONG > long_max)))

kable(tibble(
  total_rows = n_total,
  invalid_coords = nrow(bad_coords),
  pct_invalid_coords = round(100 * nrow(bad_coords) / n_total, 3)
))
```

#### 3.4 Exact duplicates (SPECIES + DATE + LAT + LONG)

```{r check-dups}
dups <- dataset %>%
  filter(!is.na(SPECIES)) %>%
  group_by(SPECIES, DATE, LAT, LONG) %>%
  filter(n() > 1) %>%
  ungroup()

kable(tibble(
  total_rows = n_total,
  exact_duplicates = nrow(dups),
  pct_duplicates = round(100 * nrow(dups) / n_total, 3)
))
```

#### 3.5 Species name sanity checks

```{r check-species}
bad_species <- dataset %>%
  filter(is.na(SPECIES) | str_detect(SPECIES, regex("S/E|S/D|SIN DATOS|\\d", ignore_case = TRUE)))

kable(tibble(
  total_rows = n_total,
  problematic_species = nrow(bad_species),
  pct_problematic_species = round(100 * nrow(bad_species) / n_total, 3)
))
```

#### 3.6 Missingness in critical fields

```{r check-na}
na_summary <- dataset %>%
  summarise(
    pct_na_species = mean(is.na(SPECIES)) * 100,
    pct_na_date = mean(is.na(DATE2)) * 100,
    pct_na_coords = mean(is.na(LAT) | is.na(LONG)) * 100,
    pct_na_georef = ifelse("GEOREFERENCING_TYPE" %in% names(.), mean(is.na(GEOREFERENCING_TYPE)) * 100, NA_real_)
  ) %>%
  mutate(across(everything(), ~ round(., 3)))

kable(na_summary)
```

---

### 4 Visualizations

#### 4.1 Uncertainty histogram (if available)

```{r plot-uncertainty, fig.width=7, fig.height=4}
if ("GEOREFERENCING_UNCERTAINTY_M" %in% names(dataset)) {
  dataset %>%
    filter(!is.na(GEOREFERENCING_UNCERTAINTY_M)) %>%
    mutate(GU = as.numeric(GEOREFERENCING_UNCERTAINTY_M)) %>%
    ggplot(aes(x = GU)) +
    geom_histogram(bins = 40, fill = "#2b8cbe", color = "white") +
    labs(x = "Georeferencing uncertainty (m)", y = "Frequency", title = "Distribution of spatial uncertainty")
} else {
  plot.new()
  text(0.5, 0.5, "Field GEOREFERENCING_UNCERTAINTY_M not available", cex = 1.1)
}
```

#### 4.2 Simple spatial scatter of valid points

```{r map-points, fig.width=7, fig.height=5}
valid_points <- dataset %>% filter(!is.na(LAT) & !is.na(LONG))
if (nrow(valid_points) > 0) {
mapgg  = ggplot(valid_points, aes(x = LONG, y = LAT)) +
    geom_point(alpha = 0.4, size = 0.8, color = "#e34a33") +
    borders("world", xlim = c(-65, -50), ylim = c(-30, -15), colour = "gray80") +
    coord_quickmap() +
    labs(x = "Longitude", y = "Latitude", title = "Spatial distribution of valid records") 
  ggplotly(mapgg)
} else {
  plot.new()
  text(0.5, 0.5, "No valid coordinate points to map", cex = 1.1)
}
```

### 4.3 densidad 


```{r map-points, fig.width=7, fig.height=5}
# Make a SpatialPoints object
occs = dataset |> dplyr::select(LONG, LAT)|>
  filter(!is.na(LAT) & !is.na(LONG))

if (nrow(occs) > 0) {
#summary(occs)

fig <- occs
fig <- fig %>%
  plot_ly(
    type = 'densitymapbox',
    lat = ~LAT,
    lon = ~LONG,
    coloraxis = 'coloraxis',
    radius = 10) 
fig <- fig %>%
  layout(mapbox= list(
    style = "open-street-map",
    zoom = 4,
    center = list(lon = -58 ,lat= -25)
  ))

fig
} else {
  plot.new()
  text(0.5, 0.5, "No valid coordinate points to map", cex = 1.1)
}

```

---


### 5 Examples of problematic records

```{r examples}
examples <- bind_rows(
  head(bad_dates, 10),
  head(bad_coords, 10),
  head(bad_species, 10),
  head(dups, 10)
) %>%
  distinct() %>%
  slice_head(n = 50)

if (!dir.exists("../logs")) dir.create("../logs", recursive = TRUE)
readr::write_csv(examples, "../logs/qa_examples.csv")
kable(head(examples, 10))
```

---

### 6 Summary of findings and recommended actions

```{r summary}
qa_summary <- tibble(
  total_rows = n_total,
  invalid_dates = nrow(bad_dates),
  invalid_coords = nrow(bad_coords),
  exact_duplicates = nrow(dups),
  problematic_species = nrow(bad_species),
  pct_na_species = na_summary$pct_na_species,
  pct_na_date = na_summary$pct_na_date,
  pct_na_coords = na_summary$pct_na_coords
)

if (!dir.exists("../reports")) dir.create("../reports", recursive = TRUE)
readr::write_csv(qa_summary, "../reports/qa_summary.csv")
kable(qa_summary)
```

**Priority actions**
- Correct transcription errors and sign mistakes in coordinates using `DATA_SOURCE` for traceability.
- Normalize species names against a taxonomic dictionary and record mappings in the processing log.
- Fix or mark invalid dates as NA and document decisions in `processing_log.csv`.
- Resolve duplicates by retaining the record with the most precise georeference or most metadata.

---

### 7 QA metadata and reproducibility

```{r metadata}
qa_meta <- tibble(
  report_generated = Sys.time(),
  dataset_path = dataset_path,
  dataset_rows = n_total,
  qa_summary_path = "../reports/qa_summary.csv",
  qa_examples_path = "../logs/qa_examples.csv",
  operator = Sys.getenv("USER", unset = "unknown")
)

readr::write_csv(qa_meta, "../logs/qa_report_metadata.csv")
session_info <- capture.output(sessionInfo())
readr::write_lines(session_info, "../logs/qa_session_info.txt")
```




